Rheology of Cubic Blue Phases

### 2013-06-06 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2000 40876441-3496-0410-a613-a2fca35b2ae2

Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
oriented along the vorticity direction. The direction of the flow
in each of the bands depends on the helicity of the underlying cholesteric
liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
patterns for left-handed and right-handed BPs are one the mirror image of
the other. 
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.

Table \ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. Therefore the averaged flow along the vorticity
direction, leading to network motion, is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative (i.e. the fluid moves but the network as a whole 
moves much less).

@@

Fig. \ref{bp1-rheo} shows the ratio between the 
apparent viscosity $\eta_{app}$ and $\eta$ versus time and 
strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
shows that there is more shear thinning in BPI than in BPII, for
the shear rate range we considered. 
Through much of the flow range we studied,
the average shear stress versus shear rate can be fitted reasonably 
well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.

For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
response versus time, and in the network morphology -- we label this 
regime BPI-2. 
The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous. We refer to this 
flow regime as BPI-1. 
The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value -- close to this breakup point the oscillations in the 
shear stress show an increasingly complex
pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
In regime BPI-3 ($6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$) 
we observe that the BPI network transforms under shear into
a  quasi two-dimensional state consisting of
double-twist rolls.
Finally, for shear rates such that
$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
configuration with the helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{cholflow}).

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
and strain (bottom).}
\label{bp1-rheo}
\end{figure}

@@

%%%  %%%%%%%%%%%%%
\ifold{The negative branch in the shear stress is connected
to a local maximum followed by a local minimum in the 
average free energy. Hence, the region of negative stress can be 
regarded as a phase during which elastic forces exert a 
'pull' on the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.}\fi

@@

Finding the recurrence period for BPI is not simple, as
this depends subtly on the applied shear rate. The analysis is complex for
a number of reasons. 
First, the recurrence period for the stress is not always the same as
that of the network configuration. 
For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
 stress cycle to complete, but $4/\gd$ for the network to recur --
this is because the configurations at times $t$ and $t+2/\gd$ are 
mirror-images. This is by no means generic though: for example
at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
stress cycle to complete, but $16/\gd$ for the network to recur. 
For some higher shear rates, the recurrence period of stress
and network conformations are instead the same.
Second, there exist isolated values (or perhaps small ranges) of shear
rates for which the recurrence period is much longer and far more irregular,
with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
Taken together, all this evidence very much suggests that 
the BPI-2 range includes at least some instances of ``rheochaos''~\cite{rheochaos,Cates:2002}.

@@

\begin{table*}[htpb]
\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
\hline
& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI & & & & $\times 10^5$\\
\hline
BPI-1 (AN) &0.24 &0.02 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &0.49 &0.04 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &0.98 &0.08 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &1.95 &0.17 & 0.1388 & & & & & & & & & \\
%BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
%BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
%BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
%BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
\hline
BPI-2 (PRC) &3.91 &0.33 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &7.81 &0.67 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &15.63 &1.33 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &15.63 &1.33 & -0.1388 & & & & & & & & & \\
BPI-2 (PRC) &23.44 &2.00 & 0.1388 & & & & & & & & & \\
%BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
%BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
%BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
%BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
%BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
\hline
BPI-3 (AMS) &31.25 &2.66 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &46.87 &4.00 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &54.69 &4.67 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &62.50 &5.33 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &78.13 &6.01 & 0.1388 & & & & & & & & & \\
%BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
\hline
BPI-4 (GJ) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
BPI-4 (GJ) &125.0 &10.66 & 0.1388 & & & & & & & & & \\
BPI-4 (GJ) &187.5 &16.00 & 0.1388 & & & & & & & & & \\
%GJT (BPI-4)& 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
\hline
BPI-5 (FAN) & 250.0 &21.33 & 0.1388 & & & & & & & & & \\
BPI-5 (FAN) & 375.0 &32.00 & 0.1388 & & & & & & & & & \\
%FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
\hline

>>>>

Characteristic bands are visible in Fig. \ref{bp2-1-velo}, which are 
oriented along the vorticity direction. The direction of the flow
in each of the bands depends on the helicity of the underlying cholesteric
liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
patterns for left-handed and right-handed BPs are one the mirror image of
the other. 
Further quantitative evidence for a direct link between the sense of motion 
and the helicity can be gained by time-averaging over individual cycles.

Table~\ref{tab1} in the Appendix 
gives minima, maxima, averages and standard deviations 
of the velocity components.
All values for the two runs with inverted helicity are identical apart 
from a change of sign in the $z$-components.
There is only a slight imbalance in magnitude 
between the maximum and minimum velocities 
along the $z$-direction. \rev{Therefore the averaged flow along the vorticity
direction is very small, and much smaller than
the local velocity flow. This suggests that the
flow of the network along the vorticity 
direction is permeative. This claim is further supported by the characteristic 
velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
whereas the network as a whole moves only in one direction.}

@@

\rev{Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
shear rates as in Fig.~\ref{bp2-flowcurve}. 
The plot shows that there is more shear thinning in BPI than in BPII, for
the shear rate range we considered.}

For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
response versus time, and in the network morphology -- we label this 
regime BPI-2. 
\rev{This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.}

The BPI-2 regime now does not persist to arbitrarily low shear: there the
oscillations become irregular and seemingly chaotic, and after a few
cycles the disclination pattern becomes amorphous.
\rev{For these very low shear rates the flow curve in Fig. \ref{bp1-flowcurve}
develops a horizontal branch, which suggest the existence of a yield stress.} 
We refer to this flow regime as BPI-1.

The BPI-2 regime is also unstable if the shear is increased {\em above} 
a critical value 
\rev{in the region of Ericksen numbers $Er \simeq O(1)$
where viscous and elastic forces are of the same magnitude. 
Close to this breakup point the network seems to become
increasingly sensitive towards small perturbations in the flow field
that cause irregularities in the order parameter structure.
These distortions couple back to the flow and form a possible
route towards "rheochaos"~\cite{rheochaos,Cates:2002}.}

We also observe that

regime BPI-3 ($3.125\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $2.67 \lesssim {\it Er} \lesssim 6.01$)

Finally, for shear rates such that
$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
configuration with the helical axis along the gradient direction as 
reported above for BPII-2 (see
section \ref{cholflow}).

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
and strain (bottom).}
\label{bp1-appvisc}
\end{figure}

@@

@@

Finding the recurrence period for BPI is not simple, as
this depends subtly on the applied shear rate. The analysis is complex for
a number of reasons. 
First, the recurrence period for the stress is not always the same as
that of the network configuration. 
For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
 stress cycle to complete, but $4/\gd$ for the network to recur --
this is because the configurations at times $t$ and $t+2/\gd$ are 
mirror-images. This is by no means generic though: for example
at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
stress cycle to complete, but $16/\gd$ for the network to recur. 
For some higher shear rates, the recurrence period of stress
and network conformations are instead the same.
Second, there exist isolated values (or perhaps small ranges) of shear
rates for which the recurrence period is much longer and far more irregular,
with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
Taken together, all this evidence very much suggests that 
the BPI-2 range includes at least some instances of "rheochaos"~\cite{rheochaos,Cates:2002}.

@@

\begin{table*}[htpb]
\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
\hline
& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI & & & & $\times 10^5$\\
\hline
BPI-1 (AN) &0.24 &0.02 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &0.49 &0.04 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &0.98 &0.08 & 0.1388 & & & & & & & & & \\
BPI-1 (AN) &1.95 &0.17 & 0.1388 & & & & & & & & & \\
%BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
%BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
%BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
%BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
\hline
BPI-2 (PRC) &3.91 &0.33 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &7.81 &0.67 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &15.63 &1.33 & 0.1388 & & & & & & & & & \\
BPI-2 (PRC) &15.63 &1.33 & -0.1388 & & & & & & & & & \\
BPI-2 (PRC) &23.44 &2.00 & 0.1388 & & & & & & & & & \\
%BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
%BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
%BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
%BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
%BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
\hline
BPI-3 (AMS) &31.25 &2.67 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &46.87 &4.00 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &54.69 &4.67 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &62.50 &5.33 & 0.1388 & & & & & & & & & \\
BPI-3 (AMS) &78.13 &6.01 & 0.1388 & & & & & & & & & \\
%BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
\hline
BPI-4 (GJ) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
BPI-4 (GJ) &125.0 &10.67 & 0.1388 & & & & & & & & & \\
BPI-4 (GJ) &187.5 &16.00 & 0.1388 & & & & & & & & & \\
%GJT (BPI-4)& 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
\hline
BPI-5 (FAN) & 250.0 &21.33 & 0.1388 & & & & & & & & & \\
BPI-5 (FAN) & 375.0 &32.00 & 0.1388 & & & & & & & & & \\
%FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
\hline
